The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-05-08 paper.qmd
cleaned up definition secion parts

# What is Uncertainty?
Uncertainty visualisation is directed towards communication - to the general public when disasters may hit, to decision makers in control centers, and a whole host of other public-facing tasks - needing an understanding of what it is when the term "uncertainty" is used.

Definitions in the literature appear seemingly view it to be encompassing and include everything the general public might consider to be uncertain. Some works (*XXX eg*) focus more narrowly on specific terms defined mathematically, such as probability, variance, error, or precision. Others (*XXX eg*) include broader loosely related elements, such as missing values. Several attempts to address this gap are made in the papers describing taxonomies.

## Definitions in taxonomies
Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the  attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc. In their own way, each of these taxonomies show an aspect of uncertainty that an author felt was important to differentiate. @utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy that was boldly referred to as a "definition of uncertainty". While taxonomies can be helpful to map out the space of "things that we consider to be related to uncertainty", a taxonomy is *not* a precise definition that would allow statisticians to estimate the uncertainty in their projects. The ramifications of this is seen throughout the literature, as not understanding how to calculate uncertainty is one of the leading reasons cited by visualisation authors when explaining why they don't include it in their visualisation [@Hullman2020a].

The task of providing an encapsulating mathematical definition of uncertainty is far beyond the scope of this work, however we will discuss an important but commonly misunderstood feature of uncertainty; its relationship to inference.

## Defining uncertainty with respect to inference
What exactly is uncertainty, then? If we were to consider making this overarching definition, what would it need in order to be "encapsulating"? Well, let us consider what might *not* be considered uncertain in order to understand this concept a little better.

@Otsuka2023 spends the first chapter of his book discussing the place of descriptive statistics in the philosophy of the field and in doing so, highlights an interesting connection between inference and uncertainty. Descriptive statistics simply describe our sample as it is and summarises large data down into an easy to swallow format. Descriptive statistics are not seen as the primary goal of modern statistics, however this was not always the case. Around the 19th century in England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson) and practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations, therefore anything that refers to the unobservable (such as new observations or population statistics) is not true science [@Otsuka2023]. By its very nature, descriptive statistics *cannot* be used to make inferences about the data because it simply exists to summaries the data, to use it to make statements about *new* data is incorrect usage. In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature" (i.e. that unobserved phenomena should be similar to observed phenomena) [@Otsuka2023]. This subtle shift, from descriptive statistics to inferential statistics was historically shunned *due to the fact it introduced the unknowable*, or in other words, uncertainty.

This philosophical understanding of statistics highlights that descriptive statistics do not have uncertainty, however some readers may disagree with that statement. Variance and probability are typically considered stand ins for "uncertainty", it is often how we choose to measure it. Additionally because probability and variance exist in descriptive statistics, descriptive statistics *must* have uncertainty. This is not necessarily true. @Begg2014 highlight that uncertainty is related to not knowing a specific value, while variability refers to the range of values a quantity can take at different locations, times or instances. Similar distinctions have been made by other authors, including @Spiegelhalter2017 who did not directly discuss variance vs uncertainty, but did differentiate between precise random events (such as the probability associated with a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased). The variance of a sample variance can be calculated and know, therefore it is not uncertain but rather it a precise description of dispersion. If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the the variance would become relevant in our discussions of uncertainty.

The idea that inference is related to uncertainty is not uncommon in the uncertainty literature, however it is often mentioned as a *task* or *goal* dependence. As the "goal" shapes every stage of an analysis, multiple authors in the uncertainty literature have commented on the need to consider quantifying and expressing uncertainty at every stage of a project [@Kinkeldey2014; @Hullman2016; @Refsgaard2007] while others highlight the importance of specific stages. @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data. @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer. @Kale2019 discussed how the choices we make in our analysis impact our outcomes and introduce uncertainty. @Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. It is clear that the distinction between noise and signal is important at every stage in an analysis, however combining the uncertainty from every stage is near impossible [@Spiegelhalter2017]. This means that uncertainty is often discussed in isolation at multiple stages of a project, which further contributes to our fractured understanding of uncertainty.

With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline. Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference. Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts. For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, @Meng2014 mathematically defined the variance introduced to a model by the array of model choices, and information theory tries to quantify uncertainty using the idea of entropy. These existing methods are not thorough enough for an analyst to understand what causes uncertainty, and quantify it for communication.

## Mistakes made when we misunderstand uncertainty
The previous section established that uncertainty is a by product of the assumptions we make when we perform inference, additionally, our quantification of uncertainty is largely task dependent, therefore the uncertainty associated with *one* distribution (which we can also view as a task or goal dependence). When uncertainty visualisation authors directly stare at this concept, they almost unanimously agree with these concepts, however when these details are pushed the the periphery of a problem, we start to get research that confuses itself. In these cases, the authors are more confused about what they are doing than the participants.

There are enough papers that fail to understand these basic principals of uncertainty that this entire paper could simply be a list with explanations. Instead, we have a handful of highlights.

@Hofman2020 performed an experiment where some participants were shown a sampling distribution of the mean, and others were shown a prediction interval around the mean, for the effectiveness of a particular treatment. The authors reported that the group that was shown the sampling distribution were willing to pay more and ascribed a higher likelihood to the belief that the treatment was more effective than a control, than the group that was shown a prediction interval. A prediction interval indicates the uncertainty associated with treatment *for a particular person* and the sampling interval shown the uncertainty associated with the *average effectiveness of the treatment*. It is already established practice in medicine to use a wide range of metrics to communicate different risks to best communicate the relevant information to a patient for a particular decision [@Spiegelhalter2017]. The perceived importance of this work by @Hofman2020 hinges on a misunderstanding that the prediction and confidence interval *are* interchangeable despite performing inference on *very different* latent variables. This misunderstanding is so widespread that this finding, that shows the confidence and prediction intervals are *not* interchangeable, is considered groundbreaking enough to be published in CHI.

@Boukhelifa2012 tried to quantify the strength of the intuitive connection between a line attribute called "sketchiness" and uncertainty. The authors of this study were aware that there was some kind of task dependence with respect to displaying uncertainty, so they tested the connection for multiple "tasks". The authors did not seem to understand the "task" dependence is likely a dependence on the inference for a particular statistic, so the "task" dependence was interpreted as *context* dependence. @fig-sketchy depicts the six scenarios that participants were shown and asked to interpret what they believe the squiggly line indicates. The authors misunderstanding led to "sketchiness" being used to depict uncertainty in the categories of bar charts, in graphic networks, and in train lines, with little consideration or explanation as to *how* these things would be uncertainty. It seems a tad ridiculous for the creators of a rail network map to be "uncertain" about the existence of a train line, however we are unable to guess at what the authors of the paper believed the uncertainty would be. The most likely scenario is that the authors did not consider the specific estimate that the uncertainty was representing, which is what led to visualisations that do not make conceptual sense. Participants rightfully assumed the sketchiness therefore represented something else, such alternative options or simply ignored it.

![The graphics displayed by @Boukhelifa2012 to identify if there is an intuitive connection between sketchiness and uncertainty. These graphic were made without concern as to what (i.e. which statistic) in the image is supposed to be uncertain. This leads to the images being difficult to connect to uncertainty even if we assume tha that is what it is supposed to represent. For example, it is unclear how you can be uncertain about a train line (S5).](sketchiness.png){#fig-sketchy width=50%}

>>>>

# What is Uncertainty?
In order to visualise uncertainty, we need to know what we are trying to communicate through visualisation.  Definitions in the literature take the view that the definition should be encompassing and include everything the general public might consider to be uncertain. Some works [@Hullman2018; @Maceachren2012; @Thomson2005] focus more narrowly on specific terms defined mathematically, such as probability, variance, error, or precision. Others [@Griethe2006; @Pang1997; @Pham2009; @Boukhelifa2017] include broader loosely related elements, such as missing values. Several attempts to address this gap are made in the papers describing taxonomies.

## Definitions of Uncertainty in the Literature
Taxonomies split uncertainty based on an endless stream of ever changing boundaries, such as whether the uncertainty is due to true randomness or a lack of knowledge [@Spiegelhalter2017; @Hullman2016; @utypo], if the uncertainty is in the  attribute, spatial elements, or temporal element of the data [@Kinkeldey2014], whether the uncertainty is scientific (e.g. error) or human (e.g. disagreement among parties) [@Benjamin2018], if the uncertainty is random or systematic [@Sanyal2009], statistical or bounded [@Gschwandtnei2016; @Olston2002], recorded as accuracy or precision [@Griethe2006; @Benjamin2018], which stage of the data analysis pipeline the uncertainty comes from [@utypo], how quantifiable the uncertainty is [@Spiegelhalter2017; @utypo], etc. In their own way, each of these taxonomies show an aspect of uncertainty that an author felt was important to differentiate. @utypo identified many common themes among this wide array of taxonomies and used them to design an encapsulating taxonomy that was boldly referred to as a "definition of uncertainty".

While taxonomies can be helpful to map out the space of "things that we consider to be related to uncertainty", a taxonomy is *not* a precise definition that would allow statisticians to estimate the uncertainty in their projects. The ramifications of this is seen throughout the literature, as a survey of visualisation authors cited "not knowing how to calculate uncertainty" as one of the primary reasons they did not include it in visualisations [@Hullman2020a]. The difficulties authors have with designing a clear definition of uncertainty likely comes from the fact that these definition fail to identify precicely what it is that *creates* uncertainty. These authors have failled to identify that uncertainty is simply a by product of inference.

## Defining uncertainty with respect to inference
It is easy to see uncertainties relationship to inference when we consider what might *not* be considered uncertain rather than just trying to think about what *is*.

Descriptive statistics simply describe our sample as it is and summarizes large data down into an easy to swallow format. Descriptive statistics are not seen as the primary goal of modern statistics, however, this was not always the case. Around the 19th century in England, *positivism* was the popular philosophical approach to science (positivists included famous statisticians such as Francis Galton and Karl Pearson) and practitioners of the approach believed statistics ended with descriptive statistics as science must be based on actual experience and observations, therefore anything that refers to the unobservable (such as new observations or population statistics) is not true science [@Otsuka2023]. In order to make statements about population statistics, future values, or new observations we need to perform inference, which requires the assumption of the "uniformity of nature" (i.e. that unobserved phenomena should be similar to observed phenomena) [@Otsuka2023]. This subtle shift, from descriptive statistics to inferential statistics was shunned during the positivism era *due to the fact it introduced the unknowable*, or in other words, uncertainty.

This approach to uncertainty is embedded in the idea that descriptive statistics do not have uncertainty, which some readers may disagree with. Specifically, because it means uncertainty is *not* a latent attribute of data, but rather an attribute of a specific hypothesis or estimate. Deniers of this fact follow a consistent logical path and it is easy to identify the common mistake. We know that variance and probability are typically considered types of "uncertainty" *and* descriptive statistics can have variance and probabilities *therefore* descriptive statistics must have uncertainty. The flaw in this logic comes from the first step, assuming that the tools with which we measure uncertainty *are* uncertainty in of themselves. This confusion is common and there are many papers that spend a great deal of time clarifying the difference. @Begg2014 highlight that uncertainty is related to not knowing a specific value, while variability refers to the range of values a quantity can take at different locations, times or instances. @Spiegelhalter2017 made sure to comment on the difference between precise random events (such as the probability associated with a coin flip), and uncertainty (such as the estimated probability associated with a coin that might be biased). The variance of a sample variance can be calculated and know, therefore it is not uncertain but rather it a precise description of dispersion. If we were to discuss drawing a new observation, or estimating the true mean of a population *then* the variance would become relevant in our discussions of uncertainty.

It is clear that uncertainties dependence on inference has gone unnoticed in the literature because authors often mention the by products of this fact as a *task* or *goal* dependence. Multiple authors have commented on the need to consider quantifying and expressing uncertainty at every stage of a project as the "goal" shapes every step of the analysis [@Kinkeldey2014; @Hullman2016; @Refsgaard2007]. @Otsuka2023 suggested that the process of observing data to perform statistics is largely dependent on our goals, because the process of boiling real world entities down into probabilistic objects (or "probabilistic kind" as he puts it) depends on the relationship we seek to identify with our data. @Meng2014 commented what is kept as data and what is tossed away is determined by the motivation of an analysis and what was previously noise can be shown to become signal depending on the the question we seek to answer. @Kale2019 discussed how the choices we make in our analysis impact our outcomes and introduce uncertainty. @Carlin2023 mentions that each research question can be can be categorised as descriptive, predictive, or causal, each of which has its own appropriate statistical methods and motivation agnostic model selection leads to statistical analysis that is devoid of meaning. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. The importance of inference when discussing uncertainty is never directly acknowledge, but it is always present. It is clear that the relationship between uncertainty and inference is noticed at every stage in an analysis, however combining these uncertainties into a single "uncertainty" value is near impossible [@Spiegelhalter2017]. With this understanding it becomes clear to see why uncertainty is tied to an endless string of examples in the data analysis pipeline. Uncertainty examples include imputed data, model selection, inherent randomness, biased sampling, etc, not because these things *are* uncertainty, but because they *create* uncertainty when we perform inference.

Existing mathematical definition of uncertainty (or related concepts) have been unable to unify this range of diverse concepts. For example, @Thomson2005 suggests a mathematical formula for *examples* of uncertainty, @Meng2014 mathematically defined the variance introduced to a model by the array of model choices, and information theory tries to quantify uncertainty using the idea of entropy. These existing methods are useful for calculating uncertainty as a specific stage of the analysis, but not thorough enough for an analyst to understand the causes of uncertainty and quantify it. The task of providing an encapsulating mathematical definition of uncertainty is far beyond the scope of this work, however focusing on its relationship to inference is likely a fruitful step forward.

## Mistakes made when we misunderstand uncertainty
The previous section established that uncertainty is a by product of the assumptions we make when we perform inference. This means that "uncertainty" is not defined by our data, but rather by the *questions we ask about it*. Unfortunately, uncertainty visualisation papers are not free from this issue. Usually papers that are confused on this concept will ask a question that is impossible to answer or does not make sense. Confusion about this core concept is incredibly common and rarely noticed, so much so that this entire paper could simply be a list of papers that have this issue. Instead, we have a handful of highlights.

@Hofman2020 performed an experiment where some participants asked to judge the effectiveness of a particular treatment when shown a sampling distribution of the mean or a prediction interval around the mean. A prediction interval indicates the uncertainty associated with treatment *for a particular person* and the sampling interval shown the uncertainty associated with the *average effectiveness of the treatment*. The authors reported that the group that was shown the sampling distribution judged the treatment as more effective than those that were shown a prediction interval, very likely because the prediction interval is always wider than the confidence interval. The importance of displaying the *most relevant* statistic and associated uncertainty is already well established in medicine, where doctors are required to understand a wide range of metrics that each communicate different risks to a patient [@Spiegelhalter2017]. Therefore, this work is not designed to commuicate to the field of medicine, but rather to the visualisation community who have historically struggled with the idea of uncertainty being related to inference. The perceived importance of this work by @Hofman2020 hinges on a widespread misunderstanding that the prediction and confidence interval *are* interchangeable despite performing inference on *very different* variables.

@Boukhelifa2012 tried to quantify the strength of the intuitive connection between a line attribute called "sketchiness" and uncertainty. The authors of this study were aware that uncertainty experiments were suffering from some kind of task dependence issue (due to other papers on the topic discussing it), but interpreted this as *context* dependence rather than dependence an inferential statistic. Participants were shown the six scenarios depicted in @fig-sketchy and asked to interpret what they believed the squiggly line indicates. A failing to understand the relationship between uncertainty and inference led the authors to add "sketchiness" to aspects of the visualization that could not possibly have an inferential statistic associated with it. It seems the authors did not consider the specific estimate that the uncertainty should represent, which is what led to visualisations that do not make conceptual sense. Participants rightfully assumed the sketchiness therefore represented something else, such alternative options or simply ignored it.

![The graphics displayed by @Boukhelifa2012 to identify if there is an intuitive connection between sketchiness and uncertainty. These graphic were made without concern as to what (i.e. which statistic) in the image is supposed to be uncertain. This leads to the images being difficult to connect to uncertainty even if we assume tha that is what it is supposed to represent. For example, (S5) implies that participants should have read the plot to mean that the creators of a rail network map would be "uncertain" about the existence of a train line, something that is defies common logic.](sketchiness.png){#fig-sketchy width=50%}
